Kindly Bent to Free Us

### 2019-07-07 kindly/intro.tex
intro-examples

These systems work well if one is prepared to write programs
functionally in resource-passing style. In this style, all operations
in the resource's API take the resource as a parameter and return it
in possibly modified state\cite{DBLP:journals/jfp/AchtenP95}. In
typestate-oriented programming, they would also modify its type
\cite{DBLP:conf/oopsla/AldrichSSS09}. Functional session types
represent a popular example \cite{DBLP:journals/jfp/GayV10}.

@@

The concepts of ownership and borrowing have grown popular over time
and they form the foundation of the type system of the Rust language
\cite{rust}. Rust supports two kinds of borrows of data structures, mutable and immutable
ones. Mutable borrows enable modification of the data structure
whereas immutable borrows come only with read access. Moreover, Rust
manages the lifetime of data and makes sure that the lifetime of a
borrow is properly contained in the lifetime of its lender.

The design of Rust is geared towards programmers with a low-level
imperative programming background, like C or C++. Its management of
lifetime supports the manual way of memory management customary in
these languages very well and makes it safe. However, programmers from
a managed language background feel alienated from the lack of garbage
collected data. They would prefer a setting where automatic memory
management with garbage collection was the default, but where they
could seemlessly switch to safe, manual resource management if that
was required.

@@

\subsection{A first example}
\label{sec:first-example}
\begin{figure}[tp]
  \begin{subfigure}[t]{0.45\linewidth}
    \lstinputlisting{code/writefiles.affe}
    \caption{File API}
    \label{fig:writing-files-api}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.45\linewidth}
\begin{lstlisting}
let main () =
  let h = File.fopen "foo" in
  write &!h "Hello ";
  write &!h "world!";
  File.close h
\end{lstlisting}
    \caption{File example}
    \label{fig:writing-files-example}
  \end{subfigure}
  \caption{Writing files}
  \label{fig:writing-fiules}
\end{figure}

As a first, well-trodden example we consider a very simple API for
writing files in \cref{fig:writing-files-api}.
The type \lstinline/File.t/ is linear. A call like
\lstinline/fopen "foo"/ returns a linear handle to a newly created
file, which \emph{must} be closed later using
\lstinline/close/. Failing to do so is a static type error.
To write to the file, we must take (mutable) borrows of the handle and
pass them to the \lstinline/write/
function. \Cref{fig:writing-files-example} contains an example where
\lstinline/&!h/ stands for the mutable borrow.

@@

\begin{itemize}
\item Type system that encodes linearity and affinity with borrowing
\item Full type inference for \dots
\item Automatic inference of regions for borrows
\end{itemize}

>>>>

These systems work well if one is prepared to write programs
functionally in resource-passing style. In this style, all operations
in the resource's API take the resource as a parameter and return it
in possibly modified state~\cite{DBLP:journals/jfp/AchtenP95}. In
typestate-oriented programming, they would also modify its
type~\cite{DBLP:conf/oopsla/AldrichSSS09}. Functional session types 
represent a popular example~\cite{DBLP:journals/jfp/GayV10}.

@@

The concepts of ownership and borrowing have grown popular over time
and they form the foundation of the type system of the Rust language
\cite{rust}. Rust supports two kinds of borrows of data structures, mutable and immutable
ones. Mutable borrows enable modification of the data structure
whereas immutable borrows grant read-only access. Mutable borrows are
exclusive whereas several immutable borrows can be active at the same
time. Moreover, Rust makes sure that the lifetime of a
borrow is properly contained in the lifetime of its lender.

The design of Rust is geared towards programmers with a low-level
imperative programming background, like C or C++. Its management of
lifetimes supports the manual way of memory management customary in
these languages very well and makes it safe. However, programmers with
a background in managed languages feel alienated from the lack of garbage
collected data. They would prefer a setting where automatic memory
management with garbage collection was the default, but where they
could seemlessly switch to safe, manual resource management if that
was required.

@@

Compared to the most closely related contender in this design space, Linear Haskell
\cite{DBLP:journals/pacmpl/BernardyBNJS18}, their goals and means are
similar as both systems permit abstraction over the number of uses of
values and retain type inference, but the details are quite different.
\begin{enumerate}
\item Their multiplicities are either linear or unrestricted whereas
  we also permit affine values.
\item Our system supports mutable and immutable borrows, hence the
  need for affinity.
\item Our multiplicities are directly attached to the type of the
  value. For example, in our system the function type \lstinline/'a-{lin}>'b/
  denotes the type of a \emph{single-use function} that can be called
  just once, whereas the multiplicities in Linear Haskell choose
  between $\alpha\to\beta$ and $\alpha \multimap\beta$ where the
  latter is a function that promises to \emph{use its argument exactly once}.
\end{enumerate}
\subsection{A first example}
\label{sec:first-example}
\begin{figure}[tp]
  \begin{subfigure}[t]{0.45\linewidth}
    \lstinputlisting{code/writefiles.affe}
    \caption{File API}
    \label{fig:writing-files-api}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.45\linewidth}
\begin{lstlisting}
let main () =
  let h = File.fopen "foo" in
  File.write &!h "Hello ";
  File.write &!h "world!";
  File.close h
\end{lstlisting}
    \caption{File example}
    \label{fig:writing-files-example}
  \end{subfigure}
  \caption{Writing files}
  \label{fig:writing-fiules}
\end{figure}

As a first, well-known example we consider a simplified API for
writing files shown in \cref{fig:writing-files-api}.  It introduces a
linear abstract type \lstinline/File.t/. A call like
\lstinline/fopen "foo"/ returns a linear handle to a newly created
file, which \emph{must} be released later on through \lstinline/close/
as shown in \cref{fig:writing-files-example}. Failing to do so is a
static type error.  To write to the file, we must take a (mutable)
borrow \lstinline/&!h/ of the handle and pass it to the
\lstinline/write/ function. Mutable borrows are classified as affine:
they must not be duplicated, but there is no requirement to use
them. This affinity shows up in the annotation of the second arrow in
the type of \lstinline/write/: a partial application like
\lstinline/write &!h/ captures the affine borrow and hence the
resulting function must also be handled in an affine way. That is, it
would be an error to write
\begin{lstlisting}
  let w = File.write &!h in w "Hello "; w "world!" (* type error *)
\end{lstlisting}
as we would be using the affine closure twice.

This example demonstrates two features of our system.
\begin{enumerate}
\item full type inference without annotations in user code
\item types carry multiplicity annotations in the form of kinds
\end{enumerate}

In general, kinds can be polymorphic and constrained. Function
application and composition are the archetypical
examples for functions exploiting that feature.
\begin{lstlisting}
let app f x = f x
# app : ('a -{'k}> 'b) -> ('a -{'k}> 'b)
\end{lstlisting}
The reading of the type inferred by our system is straightforward. If
\lstinline/f/ is a \lstinline/'k/-restricted function, then so is
\lstinline/app f/. The multiplicities of \lstinline/'a/ and
\lstinline/'b/ play no role. As usual in ML-like languages, we
implicitly assume prenex quantification by
$\forall\kappa\forall\alpha\forall\beta$. In reality, the
type checker also quantifies over the kinds of $\alpha$ and $\beta$,
hence the full truth is
$\forall\kappa\kappa_1\kappa_2\forall(\alpha:\kappa_1)\forall(\beta:\kappa_2)$,
which is only revealed if necessary for understanding the type. 
\begin{lstlisting}
let compose f g x = f (g x) 
# compose : ('k <= 'k_1) => ('b -{'k}> 'c) -> ('a -{'k_1}> 'b) -{'k}> ('a -{'k_1}> 'c)
\end{lstlisting}
Like in \lstinline/app/, the multiplicities of the type variables
\lstinline/'a,'b,'c/ do not matter. However, the multiplicity
\lstinline/'k/ of \lstinline/f/ reappears on the second to last arrow
because \lstinline/compose f/ is a closure that inherits
\lstinline/f/'s multiplicity. The multiplicities of \lstinline/g/  and
\lstinline/f/ both influence the multiplicity of the last arrow, so
that we would expect its annotation to be the least upper bound
$\kappa \sqcup \kappa_1$. Thanks to subsumption of multiplicities, it
is sufficient to assume $\kappa \le \kappa_1$ and \lstinline/g/'s
actual multiplicity gets subsumed to $\kappa_1$. This constraint
simplification technique is built into our type inference algorithm. Again, the printout
of the type scheme only mentions the required constraint
$\kappa\le\kappa_1$ and omits the prenex quantification over $\kappa,
\kappa_1$ as well as the kinds of \lstinline/'a,'b,'c/.

@@

\begin{itemize}
\item Type system that encodes linearity and affinity with borrowing
\item Full type inference for \dots
\item Constraint simplification
\item Automatic inference of regions for borrows
\end{itemize}
